### 2021-03-03 ch/1-introduction.tex
Fix bold math
Finally add the third paper

\chapter{Introduction}\label{ch:introduction}
\section{Problem statement}
In order for \ac{ML} systems to be used more widely,
they have to become more trustworthy.
The susceptibility of \acp{NN} to adversarial attacks has been well documented, but there are other problems as well.
This includes their opaqueness and their general tendency to take shortcuts;
leading to situations where neural networks do not do `what we meant', but just what they were explicitly told to do.

@@

In the setup for \emph{label bias},
there is additionally a \emph{sensitive attribute} $s$ associated with each input.
This sensitive attribute usually encodes membership in a demographic group, such as gender,
but more generally is a feature that should not be used to make predictions for legal or ethical reasons.
We will also refer to the set of all samples with a specific sensitive attribute as one \emph{demographic group}.
In our examples, $s$ and $y$ are binary variables, but this does not have to be the case.
The label bias now manifests itself in a specific way:
depending on $s$, labels are either flipped from $y=0$ to $y=1$ or vice versa,
\ie, there is an error in the labels which is correlated with the sensitive attribute.

For \emph{sampling bias},
there is also a special attribute $s$, but it does not necessarily have to denote a \emph{sensitive} attribute;
it can more generally be a \emph{spurious} variable which is correlated with the class label $y$ in the training set,
but is not truly predictive of $y$ in the general case.
Sampling bias then means that the training set is not uniformly sampled from the underlying distribution,
instead, sampling depends on $s$ and $y$;
\eg, there might be almost no samples of $s=0$ and $y=1$ in the training set.
The effect is that \ac{ML} models are tempted to take shortcuts and use $s$ as a shorthand for $y$.

@@

The fairness definition that most fits the described setup of label bias and sampling bias is \acf{DP},
also called statistical parity or independence.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them the difference and the ratio of the terms on the two sides of the equation.

@@

(MISSING HERE: ``a declaration confirming [the candidate's] contribution to each paper,
especially in cases where the co-author is a supervisor''.)

>>>>

\chapter{Introduction}\label{ch:introduction}
\section{Problem statement}
In order for \ac{ML} systems to be used more widely,
they have to become more trustworthy \citep{hleg2019ethics}.
The susceptibility of deep \acp{NN} to adversarial attacks has been well documented, but there are other problems as well.
This includes their opaqueness and their general tendency to take shortcuts;
leading to situations where neural networks do not do `what we meant', but just what they were explicitly told to do.

@@

In all cases, there is a special attribute $s$ associated with each input.
This attribute can have different meanings:
In the setting of \emph{label bias}, $s$ usually encodes membership in a demographic group, such as gender,
but more generally is a feature that should not be used to make predictions for legal or ethical reasons.
(See below for other possible meanings of $s$.)
In this setting, we call $s$ the \emph{sensitive attribute} (because it carries sensitive information).
We will often refer to the set of all samples with a specific sensitive attribute as one \emph{demographic group}.
In most of the examples, $s$ and $y$ are binary variables, but this does not have to be the case.
The label bias is related to $s$ in a very specific way:
depending on the value of $s$, labels are either flipped from $y=0$ to $y=1$ or vice versa,
\ie, there is an error in the labels which is correlated with the sensitive attribute.

For \emph{sampling bias},
there is also a special attribute $s$, but it does not necessarily have to denote a \emph{sensitive} attribute;
it can more generally be a \emph{spurious} variable which is correlated with the class label $y$ in the training set,
but is not truly predictive of $y$ in the general case.
Or, it can refer to natural \emph{subgroups} of the $y$ classes.
Sampling bias then means that the training set is not uniformly sampled from the underlying distribution:
Instead, sampling depends on $s$ and $y$;
\eg, there might be almost no samples of $s=0$ and $y=1$ in the training set.
The effect is that \ac{ML} models are tempted to take shortcuts and use $s$ as a shorthand for $y$.

@@

The fairness definition that best fits the described setup of label bias and sampling bias is \acf{DP},
also called statistical parity or independence.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them the difference and the ratio of the terms on the two sides of the equation.

@@
